<head>
  <link rel="stylesheet" type="text/css" href="/custom.css">
</head>

---
#title: "What Is Beautiful Is Good: Visualizing the Role of Visual Stereotypes in the Workplace"
---


# **Job Suitability Score Distributions Across Industries and Roles**

## **Introduction**
This visualization examines how **job suitability scores** vary **across industries and job roles**. It aims to explore whether **different roles (client-facing vs. back-office) receive systematically different ratings within Finance and Retail industries**.  

The data comes from **participant ratings** of job suitability for individuals in different roles, allowing us to analyze **whether industry and role type influence perceived job suitability**.

---


```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
#install.packages("plotly")
library(plotly)

# set default theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
# set default figure parameters for knitr
knitr::opts_chunk$set(
	fig.asp = 0.618,
	fig.height = 12,
	fig.retina = 2,
	fig.width = 17,
	dpi = 150,
	include = TRUE,
	out.width = "80%"
)

# Load actual data
data <- read.csv("org_thesis_data.csv")
```

```{r include=FALSE}
# Create Industry and Job Role categories
data <- data %>%
  mutate(
    Industry = case_when(
      condition %in% c("Financial Advisor", "Risk Analyst") ~ "Finance",
      condition %in% c("Retail Sales Associate", "Inventory Manager") ~ "Retail",
      TRUE ~ NA_character_
    ),
    JobRole = case_when(
      condition %in% c("Financial Advisor", "Retail Sales Associate") ~ "Client-facing",
      condition %in% c("Risk Analyst", "Inventory Manager") ~ "Back-office",
      TRUE ~ NA_character_
    )
  )

# Compute mean suitability scores for each Job Role within Industry
mean_data <- data %>%
  group_by(Industry, JobRole) %>%
  summarise(mean_score = mean(raw_score))

# Compute max density for each Job Role within Industry
density_data <- data %>%
  group_by(Industry, JobRole) %>%
  summarise(max_density = max(density(raw_score)$y))

# Merge max density with mean data to determine label position
mean_data <- mean_data %>%
  left_join(density_data, by = c("Industry", "JobRole")) %>%
  mutate(label_y = max_density * 0.9)  # Adjust label height



```

```{r include=FALSE}
# Generate Industry-Faceted Density Plot with Mean Lines
P_final1 <- ggplot(data, aes(x = raw_score, fill = JobRole, color = JobRole)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ Industry, ncol = 1) +  # Facet by Industry
  labs(
    title = "Mean Suitability Score Distributions by Role and Industry",
    subtitle = "Each data point represents the mean job suitability score assigned to a hyper-realistic image for a specific job title.",
    x = "Mean Suitability Score Per Stimulus",
    y = "Density"
  ) +
  scale_fill_manual(values = c("Client-facing" = "#009E73", "Back-office" = "#CC79A7")) +
  scale_color_manual(values = c("Client-facing" = "#009E73", "Back-office" = "#CC79A7")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 14, face = "bold")
  ) +
  # Grey line to cover the x-axis
  geom_hline(yintercept = 0, color = "grey", size = 1) +
  # Add vertical mean lines
  geom_vline(data = mean_data, aes(xintercept = mean_score, color = JobRole),
             linetype = "dashed", size = 1) +
  # Add mean labels dynamically positioned
  geom_text(
    data = mean_data,
    aes(x = ifelse(JobRole == "Client-facing", mean_score + 7, mean_score - 7),
        y = label_y + 0.008, 
        label = paste0("Mean: ", round(mean_score, 2))),
    inherit.aes = FALSE,
    size = 4, color = "black"
  ) +
  # Change legend name from "JobRole" to "Job Role"
  guides(fill = guide_legend(title = "Job Role"), color = "none")



```




```{r, echo=FALSE}
# Convert to Plotly with Proper Layout
ggplotly(P_final1, width = 800, height = 500) %>% 
  layout(
    title = list(
      text = "Mean Suitability Score Distributions by Role and Industry",  
      font = list(size = 16)  
    ),
    legend = list(
      orientation = "h",  
      x = 0.5,  
      y = -0.2,  
      xanchor = "center", 
      yanchor = "top",
      font = list(size = 12)  
    )
  )
```











## **Understanding the Visualization**
The density plot below shows the **distribution of mean job suitability scores** for different job roles across two industries.  

- **X-axis**: Mean suitability score rated per stimulus.  
- **Y-axis**: Density, representing the distribution of ratings.  
- **Facets**: The plot is split into **Finance** (top panel) and **Retail** (bottom panel) to highlight industry-specific differences.  
- **Colors**:  
  - **Green (Client-facing roles)**: Financial Advisor & Retail Sales Associate  
  - **Pink (Back-office roles)**: Risk Analyst & Inventory Manager  
- **Dashed vertical lines**: Represent the **mean suitability score** for each role, with **numerical labels** indicating their values.

---

## **Key Findings**
### **1. Finance Industry (Top Panel)**
- **Client-facing (Mean = 55.97) vs. Back-office (Mean = 55.25)**  
- The distributions are **highly overlapping**, suggesting that suitability scores for **Finance** roles are **consistent regardless of whether the job is client-facing or back-office**.
- This indicates that in the **Finance industry**, there may be **less appearance-based bias** in suitability judgments.

### **2. Retail Industry (Bottom Panel)**
- **Client-facing (Mean = 61.4) vs. Back-office (Mean = 56.12)**  
- Unlike in Finance, a **clearer separation** emerges, with **Client-facing roles receiving higher suitability scores** than Back-office roles.  
- This suggests that in **Retail**, people may perceive **client-facing jobs as requiring different traits**, potentially reflecting **appearance-based or personality-driven stereotypes**.

---

## **Implications & Next Steps**
- **Industry differences**: **Retail** shows stronger differentiation in job suitability ratings than **Finance**.  
- **Potential biases**: This may suggest **a preference for certain visual traits** in **customer-facing roles** within the Retail sector.  
- **Further investigation**: A statistical test (e.g., t-tests, regression) will confirm whether these **differences are statistically significant**.

This research helps us better understand how **job suitability perceptions differ by industry and role type**, with potential implications for **hiring practices and workplace biases**.

---

*Further analysis is underway to explore these patterns in greater depth!* 🚀







